# Reproducible Research - Proj Assignment 1
==========================================

This makes use of data from personal activity monitoring device and
generates calculations and plots.

#### First, set global options to show the code.

```{r setoptions, echo= TRUE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

#### Read and save the data for program calculations.
- First, set the working directory to the data directory
- Then, read the data using read.csv
- use *rowsum* to calculate the total # of steps/day
- use *hist* to plot total steps/day as a histogram
```{r readdata}
setwd("~/Desktop/Rwd/ReprodblResrch/RepResData")
pa1data <- read.csv("activity.csv")
steps.day <- rowsum(pa1data$steps, pa1data$date, na.rm = TRUE)
hist(steps.day, xlab = "Total # of Steps/day", main = "Number of Steps", col = "red")
```

#### Mean and median of total # of steps/day
```{r meanmedian}
steps.day.mean <- format(mean(steps.day), nsmall = 3, big.mark = ",")
steps.day.median <- prettyNum(median(steps.day), big.mark = ",")
```

Mean and median of the total number of steps taken per day are
`r steps.day.mean` and `r steps.day.median` respectively.

#### Average daily activity pattern

```{r adap}
suppressMessages(library(dplyr))
## group data by interval and calculate mean for each interval
## and ignore any NA's in the data
intgrp <- group_by(pa1data, interval)
intmean <- summarize(intgrp, intmean = mean(steps, na.rm = TRUE))
## plot interval vs. average for each interval
with(intmean, plot(interval, intmean, type = "l", main = "Average Daily Activity Pattern", xlab = "5-minute Intervals (hhmm)", ylab = "Average Daily Steps", col = "blue"))
## calculate the max value and the interval in which that occurs
stepmean.max <- max(intmean$intmean)
stepmean.maxintrvl <- intmean$interval[intmean$intmean == stepmean.max]
stepmean.max <- format(stepmean.max, nsmall = 3, big.mark = ",")
```

The maximum number of steps in any 5-minute interval, averaged across all days, is `r stepmean.max` and this maximum was in interval `r stepmean.maxintrvl`.

#### Inputting missing values

```{r inpmissvals}
## calculate the total # of missing values (NA's) in each row
steps.na <- sum(is.na(pa1data$steps))
date.na <- sum(is.na(pa1data$date))
interval.na <- sum(is.na(pa1data$interval))
```

The total number of missing values in the data set are

- steps: `r steps.na`,
- date: `r date.na`, and
- interval: `r interval.na`.

There are missing values only in the steps column and we shall use the following steps to back-fill the missing values:

1. Merge *pa1data* with *intmean*: this will add a column with average values for each 5-min interval across all days
2. Replace "NA" values in "*steps*" column with corresponding average values
3. Remove "*intmean*" column for the new data set "*pa1datanew*" which is same as the original with the "NA" values replaced

Using the new data set, plot a histogram of total number of steps taken each day, and calculate the the mean and median total number of steps taken each day.

```{r repnas}
pa1datanew <- merge(pa1data, intmean)
pa1datanew$steps <- with(pa1datanew, ifelse(is.na(steps), intmean, steps))
pa1datanew <- select(pa1datanew, -intmean)
stepsnew.day <- rowsum(pa1datanew$steps, pa1datanew$date)
hist(stepsnew.day, xlab = "Total # of Steps/day", main = "Number of Steps (NA's Replaced)", col = "green")
stepsnew.day.mean <- format(mean(stepsnew.day), nsmall = 3, big.mark = ",")
stepsnew.day.median <- format(median(stepsnew.day), nsmall = 3, big.mark = ",")
```

Mean and median of the total number of steps taken per day using the new data set are `r stepsnew.day.mean` and `r stepsnew.day.median` respectively. As a comparison, the same values for the original data with the "NA" values were `r steps.day.mean` and `r steps.day.median` respectively. The replacement of "NA" values with the interval mean values has smoothed out the distribution and increased the mean and median values.

#### Weekday/weekend activity patterns

```{r wdaypat}
## add a new wkend column calculated as Weekend if the days are Sat or Sun
## if not, as Weekday
pa1datanew$date <- as.Date(pa1datanew$date)
pa1datanew$wkend <- as.factor(ifelse(weekdays(pa1datanew$date) %in% c("Saturday","Sunday"), "Weekend", "Weekday"))
## group date by wkend and then by interval
wdintgrp <- group_by(pa1datanew, wkend, interval)
## calculate inteval means for Weekend and Weekday
wigrpmean <- summarize(wdintgrp, wdmean = mean(steps))
## create weekend and weekday data sets using the value of wkend
wkends <- filter(wigrpmean, wkend == "Weekend")
wkdays <- filter(wigrpmean, wkend == "Weekday")
## plot weekend and weekday data as 2 plots
par(mar=c(4,4,1,1))
par(mfcol = c(2,1))
with(wkends, plot(interval, wdmean, type = "l", main = "Weekend Activity Pattern", xlab = "5-minute Intervals (hhmm)", ylab = "Average Weekend Steps", col = "red"))
with(wkdays, plot(interval, wdmean, type = "l", main = "Weekdays Activity Pattern", xlab = "5-minute Intervals (hhmm)", ylab = "Average Weekdays Steps", col = "green"))
```

The above plots show the average number of steps across all weekdays [top plot] and all weekend days [bottom plot].

##### Note: an alternate way to get the same panel plots by using ggplot2.

```{r altplot}
wigrpmean <- summarize(wdintgrp, wdmean = mean(steps))
suppressMessages(library(ggplot2))
qplot(interval, wdmean, data = wigrpmean, facets = wkend ~. , geom = "line", main = "Average Daily Activity Pattern", xlab = "5-minute Intervals", ylab = "Average Daily Steps", color = wkend)
```

